The ChEMBL database as Linked Open Data

### 2012-04-16 article.tex
Adding Wellcome Trust acknowledgement
Acknowledgement text change
Added Compound Selectivity Use Case. The SPARQL query still causing problems, so may want to include this Use Case

@@

>>>>

\subsection{Compound Selectivity}

NOTE: May want to exclude this Use Case as query has not successful run yet, keeps causing virtuoso 
to timeout or run out of memory. Currently playing with local server setup to get query to run.

Designing a molecule, which is selective to one target over another will often be considered a 
successful outcome in a drug design process. If the target in question is a protein, it is easy
to understand that as sequence identity amongst residues, which contribute to potential small 
molecule binding sites remains high so does the issue of selectivity. That is, the molecule being 
designed may also bind to the equivalent binding site in the  closely related protein 
target and may lead to undesirable consequences. The ChEMBL data model links molecules to 
targets using different activity types recorded in the literature. Using activity types, 
which act as a measure of binding affinity, such as IC50 or Ki and applying activity value cuts
offs it is possible to identify molecules which have higher binding affinity to certain targets 
compared to others. Taking this one step further it is possible to identify a set molecules, 
which have a high affinity to protein A (e.g. IC50 value < 50nM) and low affinity to protein 
B (e.g. IC50 value > 200nM).

The following SPARQL query identifies a set of molecules, which based on data curated from the 
literature and stored in the ChEMBL-RDF data model, selectively bind Human Cyclin-Dependent Kinase 2 
(UniProt: P24941) over Human Cyclin-Dependent Kinase 4 (UniProt: P11802).

NOTE: Query should return 65 molecules

\begin{tiny}
\begin{verbatim}
select distinct ?molecule WHERE 
{ 
  ?assay ?p chembl:Assay ;
  chembl:hasTarget ?target .
  ?activity ?p chembl:Activity ;
  chembl:forMolecule ?molecule ;
  chembl:onAssay ?assay ;
  chembl:type ?type ;
  chembl:standardUnits ?unit ;
  chembl:standardValue ?value;
  chembl:relation ?relation
  FILTER (
    ?type = "IC50"  &&
    ?unit = "nM"    &&
    ?value  < 50    &&
    ?relation = "="
  ) .
  ?target ?p chembl:Target;
  dc:identifier ?id
  FILTER (
    ?id = "uniprot:P24941"
  ) .
  { 
    select ?molecule WHERE 
    {
      ?assay ?p chembl:Assay ;
      chembl:hasTarget ?target .
      ?activity ?p chembl:Activity ;
      chembl:forMolecule ?molecule ;
      chembl:onAssay ?assay;
      chembl:type ?type ;
      chembl:standardUnits ?unit ;
      chembl:standardValue ?value ;
      chembl:relation ?relation
      FILTER (
        ?type = "IC50"  &&
        ?unit = "nM"    &&
        ?value  > 200   &&
        ?relation = "="
      ) .
      ?target ?p chembl:Target;
      dc:identifier ?id
      FILTER (
        ?id = "uniprot:P11802"
      ) .
    }
  }
}
\end{verbatim}
\end{tiny}

Analysis on the set of molecules returned by this query can be used to help identify small 
molecule features, which may increase target-binding specificity. For queries which link 
small molecules to targets, by traversing bioactivity data in the ChEMBL database, it is 
also important to consider the parameters associated with the assay to target mappings. 
These additional parameters include a relationship type, a multi flag (for poorly defined
targets) a complex flag (for protein complex targets) and a curation level. These different 
factors are summarized in the ChEMBL confidence score, which ranges from 9 (direct single 
protein target) to 0 (uncurated). In order to return the largest possible dataset, the 
confidence score has been ignored in this example compound selectivity use case.

@@

The data in ChEMBL is made available by funding from the Wellcome Trust [086151/Z/08/Z]
